---
title: "ST2195 Coursework"
author: '190526753'
date: '2021 12 3 '
output:
  html_document:
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set working directory}
setwd("~/Desktop/Coursework_files")
getwd()
```

## Data preparation

#### Library needed
```{r load library}
library(vioplot)
library(tidyr)
library(ggplot2) 
library(dplyr)
library(ggstatsplot)
library(lubridate)
library(ggpubr)
library(caret)
library(Metrics)
library(superml)
library(caTools)
library(ranger)
library(rpart)
library(ROSE)
```

#### Data import and preparation 

* Read all the data we will be using.
```{r read data}
airports <- read.csv("./dataverse_files/airports.csv", sep=",", header = TRUE)
carriers <- read.csv("./dataverse_files/carriers.csv", sep=",", header = TRUE)
planes <- read.csv("./dataverse_files/plane-data.csv", sep=",", header = TRUE)
variable_descriptions <- read.csv("/Users/sangbin/Desktop/Coursework_files/dataverse_files/variable-descriptions.csv",sep=",", header = TRUE)

df_03 <- read.csv("./dataverse_files/2003.csv.bz2", sep=",", header = TRUE)
df_04 <- read.csv("./dataverse_files/2004.csv.bz2", sep=",", header = TRUE)
df_05 <- read.csv("./dataverse_files/2005.csv.bz2", sep=",", header = TRUE)

```
<br/>

* Concatenate all the flight detail dataset into one dataset.
```{r caocatenate datasets}
df <- rbind(df_03, df_04, df_05)
```
<br/>

* Create full date column
```{r create date}
df$Date <- make_date(df$Year,df$Month,df$DayofMonth)
```
<br/>

* We are able to identify that if the flight is cancelled or diverted, there is no delay occurred. Hence we will filter out the cancelled and diverted flights for questions 1 to 4 just for now.

<br/>
```{r df_copy}
df_copy <- df
df_copy <- subset(df_copy, Cancelled == 0 & Diverted == 0)
df_copy <- subset(df_copy, select=-c(CancellationCode))

```
<br/>

* Make new columns to indicate delay of arrival and departure. 0 = on time , 1 = delayed

```{r delay column creation}
df_copy$DelayedArr <- 0
df_copy$DelayedDep <- 0
```
<br/>

* According to Bureau of Transportation statistics, a flight is counted as "on time" if the flight
operated less than 15 minutes later than the scheduled time shown in the carriers' Computerized Reservations Systems (CRS).Thus, On-time means a flight that arrives less than 15 minutes after its announced arrival time.

```{r delayed flights}
df_copy$DelayedArr[df_copy$ArrDelay > 15] <- 1
df_copy$DelayedDep[df_copy$DepDelay > 15] <- 1
```
<br/>

* Missing data (or Not a Number data) tells us that these features are not the cause of it. Hence convert NaN values into 0

```{r NA values}
df_copy$CarrierDelay <- df_copy$CarrierDelay %>% replace_na(0)
df_copy$WeatherDelay <- df_copy$WeatherDelay %>% replace_na(0)
df_copy$NASDelay <- df_copy$NASDelay %>% replace_na(0)
df_copy$SecurityDelay <- df_copy$SecurityDelay %>% replace_na(0)
df_copy$LateAircraftDelay <- df_copy$LateAircraftDelay %>% replace_na(0)
```
<br/>

* We can identify that there are no missing values except "ArrDelay" and "ActualElapsedTime".
Hence, drop na values
```{r}
summary(df_copy)
df_copy <- df_copy %>% drop_na(ArrDelay)
```
<br/>

* convert concatenated hours and minutes into separate columns Hour

```{r}
df_copy$DepHour = df_copy$DepTime %/% 100L
df_copy$ArrHour = df_copy$ArrTime %/% 100L
df_copy$CRSDepHour = df_copy$CRSDepTime %/% 100L
df_copy$CRSArrHour = df_copy$CRSArrTime %/% 100L
```
<br/>

* Convert Hour's data type numeric to character.
```{r}
df_copy$DepHour <- as.character(df_copy$DepHour)
df_copy$ArrHour <- as.character(df_copy$ArrHour)
df_copy$CRSDepHour <- as.character(df_copy$CRSDepHour)
df_copy$CRSArrHour <- as.character(df_copy$CRSArrHour)
```
<br/>

* Convert the time into 24-hour notation.
```{r}
df_copy$DepHour = with(df_copy, ifelse(DepHour %in% c("24","0") , "00",
                                       ifelse(DepHour %in% c("25","1") , "01",
                                              ifelse(DepHour %in% c("26","2") , "02",
                                                     ifelse(DepHour %in% c("27","3") , "03",
                                                            ifelse(DepHour %in% c("28","4") , "04",
                                                                   ifelse(DepHour %in% c("29","5") , "05",
                                                                          ifelse(DepHour %in% c("6") , "06",
                                                                                 ifelse(DepHour %in% c("7") , "07",
                                                                                        ifelse(DepHour %in% c("8") , "08",
                                                                                               ifelse(DepHour %in% c("9") , "09",
                                                                                                      DepHour)))))))))))

df_copy$ArrHour = with(df_copy, ifelse(ArrHour %in% c("24","0") , "00",
                                       ifelse(ArrHour %in% c("25","1") , "01",
                                              ifelse(ArrHour %in% c("26","2") , "02",
                                                     ifelse(ArrHour %in% c("27","3") , "03",
                                                            ifelse(ArrHour %in% c("28","4") , "04",
                                                                   ifelse(ArrHour %in% c("29","5") , "05",
                                                                          ifelse(ArrHour %in% c("6") , "06",
                                                                                 ifelse(ArrHour %in% c("7") , "07",
                                                                                        ifelse(ArrHour %in% c("8") , "08",
                                                                                               ifelse(ArrHour %in% c("9") , "09",
                                                                                                      ArrHour)))))))))))

df_copy$CRSDepHour = with(df_copy, ifelse(CRSDepHour %in% c("24","0") , "00",
                                       ifelse(CRSDepHour %in% c("25","1") , "01",
                                              ifelse(CRSDepHour %in% c("26","2") , "02",
                                                     ifelse(CRSDepHour %in% c("27","3") , "03",
                                                            ifelse(CRSDepHour %in% c("28","4") , "04",
                                                                   ifelse(CRSDepHour %in% c("29","5") , "05",
                                                                          ifelse(CRSDepHour %in% c("6") , "06",
                                                                                 ifelse(CRSDepHour %in% c("7") , "07",
                                                                                        ifelse(CRSDepHour %in% c("8") , "08",
                                                                                               ifelse(CRSDepHour %in% c("9") , "09",
                                                                                                      CRSDepHour)))))))))))

df_copy$CRSArrHour = with(df_copy, ifelse(CRSArrHour %in% c("24","0") , "00",
                                       ifelse(CRSArrHour %in% c("25","1") , "01",
                                              ifelse(CRSArrHour %in% c("26","2") , "02",
                                                     ifelse(CRSArrHour %in% c("27","3") , "03",
                                                            ifelse(CRSArrHour %in% c("28","4") , "04",
                                                                   ifelse(CRSArrHour %in% c("29","5") , "05",
                                                                          ifelse(CRSArrHour %in% c("6") , "06",
                                                                                 ifelse(CRSArrHour %in% c("7") , "07",
                                                                                        ifelse(CRSArrHour %in% c("8") , "08",
                                                                                               ifelse(CRSArrHour %in% c("9") , "09",
                                                                                                      CRSArrHour)))))))))))

```
<br/>

* Groupby distance to see the relationship between delays
```{r}
Distance_delay <- df_copy %>%
  group_by(Distance) %>%
  summarise(Average_ArrDelay = mean(ArrDelay))
```
<br/>

* Average arrival delay based on distance and Frequency of the flight over distance.
```{r, fig.width=14, fig.height=12}
a<- ggplot(data = Distance_delay , mapping = aes(x = Distance, y = Average_ArrDelay)) +
  geom_line(stat='identity') +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of arrival delay based on distance",x = "Distance", y ="Average ArrDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))

b<-ggplot(data = Distance_delay ,aes(x = Distance)) +
  geom_histogram(aes(y=..density..),
                 binwidth=100) +
  labs(title = "Frequency of the flight over distance",x = "Distance", y ="Frequency of the flight") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))+
  geom_density(alpha=.2, fill="white")

ggarrange(a,b, 
          labels = c("A", "B"),
          ncol = 2, nrow = 1)
```

* We can observe that shorter distanced flights seem to have more fluctuation on delays compared to long distance flights.

* We can also observe that there are more of shorter distanced flights compared to long distance flights.

<br/>

## 1) When is the best time of day, day of the week, and time of year to fly to minimize delays?

<br/>

#### Deep dive into Delayed arrival flights

```{r}
Delayed_flights <- filter(df_copy,df_copy$DelayedArr == 1)

```
<br/>

* ArrDelay Boxplot(before)
```{r}
c <- ggplot(data=Delayed_flights, aes(y=ArrDelay))+
  geom_boxplot(fill='slategrey',color='darkslategrey',width=0.3)+
  labs(title = "ArrDelay Boxplot(before)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```
<br/>

* DepDelay Boxplot(before)
```{r, fig.width=14, fig.height=12}
d <- ggplot(data=Delayed_flights, aes(y=DepDelay))+
  geom_boxplot(fill='slategrey',color='darkslategrey',width=0.3)+
  labs(title = "DepDelay Boxplot(before)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))

ggarrange(c,d, 
          labels = c("C", "D"),
          ncol = 2, nrow = 1)

```
<br/>

* Identify first, third quartile and interquartile range to identify the lower and upper whisker
```{r}
Q_ArrDelay <- quantile(Delayed_flights$ArrDelay, probs=c(.25, .75), na.rm = FALSE)
IQR_ArrDelay <- IQR(Delayed_flights$ArrDelay)
Upper_Whisker_ArrDelay <-  Q_ArrDelay[2]+1.5*IQR_ArrDelay
Lower_Whisker_ArrDelay <- Q_ArrDelay[1]-1.5*IQR_ArrDelay
Q_DepDelay <- quantile(Delayed_flights$DepDelay, probs=c(.25, .75), na.rm = FALSE)
IQR_DepDelay <- IQR(Delayed_flights$DepDelay)
Upper_Whisker_DepDelay <-  Q_DepDelay[2]+1.5*IQR_DepDelay
Lower_Whisker_DepDelay <- Q_DepDelay[1]-1.5*IQR_DepDelay
```
<br/>

* Outliers will be any points below Lower_Whisker or above Upper_Whisker hence sort them out from arrival delayed flights.

```{r}
Delayed_flights<- subset(Delayed_flights, Delayed_flights$ArrDelay > Lower_Whisker_ArrDelay & 
                           Delayed_flights$ArrDelay < Upper_Whisker_ArrDelay &
                           Delayed_flights$DepDelay > Lower_Whisker_DepDelay & 
                           Delayed_flights$DepDelay < Upper_Whisker_DepDelay)
```
<br/>

* ArrDelay Boxplot (after)
```{r}
e <- ggplot(data=Delayed_flights, aes(y=ArrDelay))+
  geom_boxplot(fill='slategrey',color='darkslategrey',width=0.3)+
  labs(title = "ArrDelay Boxplot(after)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```
<br/>

* DepDelay Boxplot (after)
```{r}
f <- ggplot(data=Delayed_flights, aes(y=DepDelay))+
  geom_boxplot(fill='slategrey',color='darkslategrey',width=0.3)+
  labs(title = "DepDelay Boxplot(after)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))

ggarrange(e,f, 
          labels = c("E", "F"),
          ncol = 2, nrow = 1)
```
<br/>

* Total number of flights per hour
```{r}
total_flights_per_DepHour <- df_copy %>%
  group_by(DepHour) %>%
  summarise(Count = sum(DelayedArr))

g <- ggplot(data = total_flights_per_DepHour, mapping = aes(x = reorder(DepHour,-Count), y = Count)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Count,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Total number of flights per hour",x = "Actual Departure Hour", y ="Number of flights") + 
  theme(plot.title = element_text(hjust = 0.5))
```
<br/>

* Number of delayed flights per hour
```{r, fig.width=14, fig.height=12}
delayed_flights_per_DepHour <- Delayed_flights %>%
  group_by(DepHour) %>%
  summarise(Count = sum(DelayedArr))

h <- ggplot(data = delayed_flights_per_DepHour, mapping = aes(x = reorder(DepHour,-Count), y = Count)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Count,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Number of delayed flights per hour",x = "Actual Departure Hour", y ="Number of delayed flights") + 
  theme(plot.title = element_text(hjust = 0.5))

ggarrange(g,h, 
          labels = c("G", "H"),
          ncol = 2, nrow = 1)
```

* We can identify that number of flights differ by hour. 

</br>

* Actual Departure Hour
```{r}
Delayed_flights_DepHour <- Delayed_flights %>%
  group_by(DepHour) %>%
  summarise(Average_DepDelay = mean(DepDelay),Average_ArrDelay = mean(ArrDelay))

i <- ggplot(data = Delayed_flights_DepHour, mapping = aes(x = DepHour, y = Average_DepDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_DepDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of delay over the hours day",x = "Actual Departure Hour", y ="Average DepDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5))

j <- ggplot(data = Delayed_flights_DepHour, mapping = aes(x = DepHour, y = Average_ArrDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of delay over the hours day",x = "Actual Departure Hour", y ="Average ArrDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5))

```
<br/>

* CRSDeparture Hour
```{r, fig.width=14, fig.height=12}
Delayed_flights_CRSDepHour <- Delayed_flights %>%
  group_by(CRSDepHour) %>%
  summarise(Average_DepDelay = mean(DepDelay),Average_ArrDelay = mean(ArrDelay))

k <- ggplot(data = Delayed_flights_CRSDepHour, mapping = aes(x = CRSDepHour, y = Average_DepDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_DepDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of delay over the hours day",x = "CRS Departure Hour", y ="Average DepDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5))


l <- ggplot(data = Delayed_flights_CRSDepHour, mapping = aes(x = CRSDepHour, y = Average_ArrDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of delay over the hours day",x = "CRS Departure Hour", y ="Average ArrDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5))

ggarrange(i,j,k,l, 
          labels = c("I", "J","K","L"),
          ncol = 2, nrow = 2)
```
<br/>

* Identify the best day of week to book a flight

```{r}
Delayed_flights_Day <- Delayed_flights %>%
  group_by(DayOfWeek) %>%
  summarise(Count = sum(DelayedArr),Average_DepDelay = mean(DepDelay),Average_ArrDelay = mean(ArrDelay))

m <-ggplot(data = Delayed_flights_Day, mapping = aes(x = reorder(wday(DayOfWeek,label = TRUE),-Average_ArrDelay), y = Average_ArrDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
   labs(title = "Average Arrival Delay by Day",x = "Day", y ="Average ArrDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5))
```
<br/>

* Number of Arrival Delay by Day
```{r, fig.width=14, fig.height=12}
n <- ggplot(data = Delayed_flights_Day, mapping = aes(x = reorder(wday(DayOfWeek,label = TRUE),-Count), y =Count)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Count,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Number of Arrival Delay by Day",x = "Day", y ="Arrival Delay Count") + 
  theme(plot.title = element_text(hjust = 0.5))

ggarrange(m,n, 
          labels = c("M", "N"),
          ncol = 2, nrow = 1)

```
<br/>

* Identify the best day of month to book a flight

```{r}
Delayed_flights_Month <- Delayed_flights %>%
  group_by(Month) %>%
  summarise(Count= sum(DelayedArr),Average_DepDelay = mean(DepDelay),Average_ArrDelay = mean(ArrDelay))

o <- ggplot(data = Delayed_flights_Month, mapping = aes(x = reorder(month(Month,label = TRUE),-Average_ArrDelay), y = Average_ArrDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Average Arrival Delay by Month",x = "Month", y ="Average ArrDelay (min)")+
  theme(plot.title = element_text(hjust = 0.5)) 
```

<br/>

* Number of Arrival Delay by Month

```{r, fig.width=14, fig.height=12}
p <- ggplot(data = Delayed_flights_Month, mapping = aes(x = reorder(month(Month,label = TRUE),-Count), y =Count)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Count,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Number of Arrival Delay by Day",x = "Month", y ="Arrival Delay Count") + 
  theme(plot.title = element_text(hjust = 0.5))

ggarrange(o,p, 
          labels = c("O", "P"),
          ncol = 2, nrow = 1)
```
<br/>

*From the plots, we can identify the best time to minimize delays.

<br/>

## 2) Do older planes suffer more delays?

* To check whether older planes suffer more delays, use delayed flight dataset from question 1 which excludes the outliers. Then, merge planes dataset which includes aircraft details into delayed flight dataset.

```{r}
plane_characteristic <- Delayed_flights 
  
plane_characteristic <- left_join(plane_characteristic, planes, by = c("TailNum"='tailnum'))%>%
  group_by(year) %>%
  summarise(Average_DepDelay = mean(DepDelay),Average_ArrDelay = mean(ArrDelay), Average_CarrierDelay = mean(CarrierDelay))
```
<br/>

* Drop non-indicated year values
```{r}
plane_characteristic <- plane_characteristic %>%
  filter(!is.na(year), plane_characteristic$year !="",plane_characteristic$year !="0000",plane_characteristic$year !="None",plane_characteristic$year !="2007")  
```
<br/>

* Plot Average delay of each engine year

<br/>

* Average Arrival Delay
```{r}
q<- ggplot(data = plane_characteristic, mapping = aes(x = year, y = Average_ArrDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of delay over the engine year of the plane",x = "Engine Year", y ="Average ArrDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```
<br/>

* Average Departure Delay
```{r}
r<- ggplot(data = plane_characteristic, mapping = aes(x = year, y = Average_DepDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of delay over the engine year of the plane",x = "Engine Year", y ="Average DepDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```
<br/>

* Average Carrier Delay
```{r, fig.width=14, fig.height=12}
s<- ggplot(data = plane_characteristic, mapping = aes(x = year, y = Average_CarrierDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of delay over the engine year of the plane",x = "Engine Year", y ="Average CarrierDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))

ggarrange(q,r,s, 
          labels = c("Q", "R", "S"),
          ncol = 2, nrow = 2)

```

* From the graph, we don't see much of fluctuation or difference between each engine year. Hence, we can assume that age of the plane does not have great impact on its delays.
<br/>

## 3) How does the number of people flying between different locations change over time?

<br/>

* copy preprocessed data and label the month of each flight
```{r}
Destination <- df_copy
Destination$season <- month(Destination$Date,label = TRUE) 
```
<br/>

* Concatenate flights origin and destination to observe its route.
```{r}
Destination$trip <-paste(Destination$Origin,Destination$Dest,sep="-")
```
<br/>

* Indicate season for each month
* December to February = Winter
* March to May = Spring
* June to August = Summer
* September to November = Autumn
```{r}
Destination$season = with(Destination, ifelse(season%in% c("Dec","Jan","Feb") , "Winter",
                                       ifelse(season %in% c("Mar","Apr","May") , "Spring",
                                              ifelse(season %in% c("Jun","Jul","Aug") , "Summer",
                                                     ifelse(season %in% c("Sep","Oct","Nov") , "Autumn",
                                                            season)))))
```

<br/>

* Merge airport dataset which includes airport details.       
```{r}
Destination <- left_join(Destination, airports, by = c("Dest"='iata'))
```
<br/>

* To identify the number of people flying between different locations over time, plot most visited trips, airport, city, and state.

* Top 10 Most Frequent Trips
```{r, fig.width=14, fig.height=12}
Destination_trip <- Destination %>%
  group_by(season,trip)%>%
  summarise(count = n())

  Destination_trip %>% 
  arrange(desc(count)) %>%
  slice(1:10) %>%
  ggplot(., mapping = aes(x = reorder(trip, -count), y = count)) +
  facet_wrap(~season)+
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(count,0), fontface = 'bold', vjust = -0.2), size = 1.5) +
  labs(title = "Top 10 Most Frequent Trips",x = "Trip", y ="Total Trip Count") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```

<br/>

* Top 10 Most Visited Airport
```{r, fig.width=14, fig.height=12}
Destination_airport <- Destination %>%
  group_by(season,airport)%>%
  summarise(count = n())

Destination_airport %>% 
  arrange(desc(count)) %>%
  slice(1:10) %>%
  ggplot(., mapping = aes(x = reorder(airport, -count), y = count)) +
  facet_wrap(~season)+
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(count,0), fontface = 'bold', vjust = -0.2), size = 1.5) +
  labs(title = "Top 10 Most Frequent airports",x = "airport", y ="Total airport Count") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```

<br/>

* Top 10 Most Visited City
```{r, fig.width=14, fig.height=12}
Destination_city <- Destination %>%
  group_by(season,city)%>%
  summarise(count = n())

  Destination_city %>% 
  arrange(desc(count)) %>%
  slice(1:10) %>%
  ggplot(., mapping = aes(x = reorder(city, -count), y = count)) +
  facet_wrap(~season)+
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(count,0), fontface = 'bold', hjust = .5,vjust = 2), size = 2) +
  labs(title = "Top 10 Most Frequent citys",x = "city", y ="Total city Count") + 
  theme(plot.title = element_text(hjust = 0.5), # title
        axis.text.x=element_text(angle=90, hjust=1, size = 8), # x axis
        axis.text.y=element_text(hjust=1, size = 8), # y axis
        strip.text = element_text(size = 8))#facet text

```
<br/>

* Top 10 Most Visited State
```{r, fig.width=14, fig.height=12}
Destination_state <- Destination %>%
  group_by(season,state)%>%
  summarise(count = n())

  Destination_state %>% 
  arrange(desc(count)) %>%
  slice(1:10) %>%
  ggplot(., mapping = aes(x = reorder(state, -count), y = count)) +
  facet_wrap(~season)+
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(count,0), fontface = 'bold', hjust = .5,vjust = 2), size = 2) +
  labs(title = "Top 10 Most Frequent citys",x = "city", y ="Total city Count") + 
  theme(plot.title = element_text(hjust = 0.5), # title
        axis.text.x=element_text(angle=90, hjust=1, size = 8), # x axis
        axis.text.y=element_text(hjust=1, size = 8), # y axis
        strip.text = element_text(size = 8))#facet text
  
```


* We divided 4 different sections to demonstrate how the number of people fly between different locations change over time. As we can see from the plot, most frequent trip happened in between Los Angeles International Airport (LAX) and San Diego International Airport (SAN) during winter compared to other seasons. This makes sense because temperatures in San Diego are sufficiently warm year-round, hence we can assume that people tend to visit San Diego during winter to get rid of the cold weather and fill the warm air. Similar finding could be applicable for other flight's route too. Moreover, we can identify that having the most frequent trip doesn’t necessarily mean it would be parallel to the most visited airport, city and state since numerous airports are correlated with lots of different city and state. Thus, "Top 10 Most Visited Airport" plot with William B Hartsfield-Atlanta Intl airport as the most visited airport located in the city of Atlanta is contrasting to "Top 10 Most Visited City" plot showing Chicago as the most visited city. Furthermore, people tend to travel more on spring and summer when the weather is ideal. For summer, it may be due to summer vacation trips. While for spring, considered as off-season compared to summer, could have fewer crowds and the prices of touring may be less compared to summer when prices soar which would then result to encourage people to travel during spring. Hence month and origin of the airport could be an important indicator for prediction model.
  
## 4) Can you detect cascading failures as delays in one airport create delays in others?

<br/>

* First, cascading delay occurs when a delay at previous airport flight cause a ripple effect in the following airport flight.

<br/>

* We used "Destination" dataset from question 3 which includes the airport details.Then filtered out flights with no "LateAircraftDelay" from "Destination" DF since we are trying to detect cascading failures as delays in one airport create delays in others.

```{r}
Airport <- filter(Destination,Destination$LateAircraftDelay > 0) 
```
<br/>

* In addition, Filter out flights with no delayed arrival and departure

```{r}
Airport <- subset(Airport, Airport$DelayedArr == 1 & Airport$DelayedDep == 1)
```
<br/>

* We will investigate flights with Tail Number "N336" and "N337"  to find the cascading failures as delays in one airport create delays in others.

```{r}
Airport_cascading_failure <- subset(Airport, Airport$TailNum %in% c("N336","N337"))
```
<br/>

* we can identify that LateAircraftDelay seems decreasing
```{r}
no_cascading_failure <- Airport_cascading_failure %>%
        filter(Date =="2004-04-12" & FlightNum == 2967)%>%
        arrange(CRSDepHour)

t <- ggplot(data = no_cascading_failure, mapping = aes(x = CRSDepHour, y = LateAircraftDelay)) +
  geom_line(group=1) +
  geom_text(mapping = aes(label = LateAircraftDelay, fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "LateAircraftDelay on 2004-04-12 for tail N336",x = "CRSDepHour", y ="LateAircraftDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5))
```

<br/>

* We can identify that LateAircraftDelay seems increasing
```{r, fig.width=14, fig.height=12}
cascading_failure <- Airport_cascading_failure %>%
  filter(Date =="2005-06-10" & TailNum == "N337")%>%
  arrange(CRSDepHour)

u <- ggplot(data = cascading_failure, mapping = aes(x = CRSDepHour, y = LateAircraftDelay)) +
  geom_line(group=1) +
  geom_text(mapping = aes(label = LateAircraftDelay, fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "LateAircraftDelay on 2005-06-10 for tail N337",x = "CRSDepHour", y ="LateAircraftDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5))

ggarrange(t,u, 
          labels = c("T", "U"),
          ncol = 2, nrow = 1)

```

* We can identify that LateAircraftDelay seems decreasing and increasing in some cases.

<br/>

* To further investigate, plot top 10 airports with longest delay.

```{r}
Longest_carrier_delay_flights <- Airport %>%
  group_by(airport) %>%
  summarise(Average_DepDelay = mean(DepDelay),Average_ArrDelay = mean(ArrDelay), Average_CarrierDelay = mean(CarrierDelay))
```
<br/>

* Top 10 airports with longest departure delay
```{r}
v <- Longest_carrier_delay_flights %>% 
  arrange(desc(Average_DepDelay)) %>%
  slice(1:10) %>%
  ggplot(., mapping = aes(x = reorder(airport, -Average_DepDelay), y =Average_DepDelay)) +
  geom_bar(stat='identity',fill="steelblue")+
  geom_text(mapping = aes(label = round(Average_DepDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Top 10 airports with longest departure delay",x = "Airport", y ="Average DepDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```
<br/>

* Top 10 airports with longest arrival delay
```{r}
w <- Longest_carrier_delay_flights %>% 
  arrange(desc(Average_ArrDelay)) %>%
  slice(1:10) %>%
  ggplot(., mapping = aes(x = reorder(airport, -Average_ArrDelay), y =Average_ArrDelay)) +
  geom_bar(stat='identity',fill="steelblue")+
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Top 10 airports with longest arrival delay",x = "Airport", y ="Average ArrDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```
<br/>

* Top 10 airports with longest carrier delay
```{r, fig.width=14, fig.height=12}
x <- Longest_carrier_delay_flights %>% 
  arrange(desc(Average_CarrierDelay)) %>%
  slice(1:10) %>%
  ggplot(., mapping = aes(x = reorder(airport, -Average_CarrierDelay), y =Average_CarrierDelay)) +
  geom_bar(stat='identity',fill="steelblue")+
  geom_text(mapping = aes(label = round(Average_CarrierDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Top 10 airports with longest carrier delay",x = "Airport", y ="Average CarrierDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))

ggarrange(v,w,x,
          labels = c("V", "W", "X"),
          ncol = 2, nrow = 2)
```

* As it gives top 10 of all different airports for each different delays, we can assume that there is a fluctuation of average delays for different airports.

<br/>

* Furthermore, check if different airlines have unusual characteristics on its delays.

<br/>

* Distribution of average delay over Airline

```{r}
Airline <- Delayed_flights 
Airline <- left_join(Airline, carriers, by = c("UniqueCarrier"='Code'))%>%
  group_by(UniqueCarrier) %>%
  summarise(Average_DepDelay = mean(DepDelay),Average_ArrDelay = mean(ArrDelay), Average_CarrierDelay = mean(CarrierDelay))
```
<br/>

* Average Arrival Delay
```{r}
y <- ggplot(data = Airline, mapping = aes(x = UniqueCarrier, y = Average_ArrDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of arrival delay over Airline",x = "Airline Code", y ="Average ArrDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```
<br/>

* Average Departure Delay
```{r}
z <- ggplot(data = Airline, mapping = aes(x = UniqueCarrier, y = Average_DepDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of departure delay over Airline",x = "Airline Code", y ="Average DepDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))
```
<br/>

* Average Carrier Delay
```{r, fig.width=14, fig.height=12}
aa<- ggplot(data = Airline, mapping = aes(x = UniqueCarrier, y = Average_CarrierDelay)) +
  geom_bar(stat='identity',fill="steelblue") +
  geom_text(mapping = aes(label = round(Average_ArrDelay,0), fontface = 'bold', vjust = -0.2), size = 2) +
  labs(title = "Distribution of carrier delay over Airline",x = "Airline Code", y ="Average CarrierDelay (min)") + 
  theme(plot.title = element_text(hjust = 0.5),axis.text.x=element_text(angle=90, hjust=1))

ggarrange(y,z,aa,
          labels = c("Y", "Z", "AA"),
          ncol = 2, nrow = 2)
```

* We observed some fluctuation of average delays between different airports and airlines. Hence, we concluded that delays could vary based on their origin of the airports and their airlines.

## 5) Use the available variables to construct a model that predicts delays.

<br/>

* Since the dataset is huge, we will be modeling using 0.1% of the original dataset.

```{r}
prediction <- df_copy %>% filter(row_number() %% 1000 == 0) %>%
  dplyr::select(c(DelayedArr,CRSDepHour,Origin,UniqueCarrier,Month,Distance))
```

<br/>

* Turn delayed arrivals into a factor
```{r}
prediction[prediction$DelayedArr == 0,]$DelayedArr <- "on-time"
prediction[prediction$DelayedArr == 1,]$DelayedArr <- "delayed"
prediction$DelayedArr <- as.factor(prediction$DelayedArr)
```
<br/>

* Encode categorical features we will be using into numerical value to allow the model to perform with it.
```{r}
lbl <- LabelEncoder$new()

prediction$CRSDepHour <- lbl$fit_transform(prediction$CRSDepHour)
prediction$Origin <- lbl$fit_transform(prediction$Origin)
prediction$UniqueCarrier <- lbl$fit_transform(prediction$UniqueCarrier)
```
<br/>

* Create a list of 80% of the rows in the prediction dataset.
```{r}
set.seed(1)
training_sample <- createDataPartition(prediction$DelayedArr,p = 0.8, list = FALSE)
```
<br/>

* Use the 80% of data for training and testing the models.
```{r}
dataset<- prediction[training_sample,]
```
<br/>

* Use the 20% of the data to validate the model.
```{r}
validation <- prediction[-training_sample,]
```

* Since our aim is to predict delayed flights, pick "DelayedArr" as independent variable(label). Pick dependent variables(features) that are available in the future flights and assumed to be influential to delays.
<br/>
<br/>

* split input and output
```{r}
output <- dataset[,1]

input <- dataset[,2:6]
```
<br/>

* Violin plot for each attribute on one image
```{r}
par(mfrow=c(1,5))
for(i in 2:6) {vioplot(prediction[,i], main=names(input)[i-1])}

```
<br/>

* We will be using 10-fold cross-validation
```{r}
control <- trainControl(method="cv", number=10)
metric <- "Accuracy"
```
<br/>

* From the table, we can observe that delayed is much less compared to on-time. This may cause class imbalance which results to bias.

```{r}
table(dataset$DelayedArr)
```

<br/>

* Since we are interested on delayed flights, we need to revise the model because on-time flight is dominated over delayed flight.

```{r}
prop.table(table(dataset$DelayedArr))

par(mfrow=c(1, 1))
barplot(prop.table(table(dataset$DelayedArr)),
        ylim = c(0, 0.9),
        main = "Class Distribution")

```

* Proportion rate of class shows 19% for delayed flights and 81% for on-time flights.

<br/>

* To make the class balanced, we will be using 3 techniques to balance the class

<br/>

* Under sampling
```{r}
set.seed(1)
undersampling <- ovun.sample(DelayedArr~., data=dataset, method = "under",N = 6082)$data
table(undersampling$DelayedArr)

```
<br/>

* Oversampling
```{r}
set.seed(1)
oversampling <- ovun.sample(DelayedArr~., data=dataset, method = "over",N = 26488)$data
table(oversampling$DelayedArr)
```
<br/>

* Both (Over & Under)
```{r}
set.seed(1)
bothsampling <- ovun.sample(DelayedArr~., data=dataset, method = "both",N = 16285, p=.5)$data
table(bothsampling$DelayedArr)
```
<br/>

* with 3 techniques, compute the model using each data and evaluate its accuracy then build decision tree models.

```{r}
tree.over <- rpart(DelayedArr~., data=oversampling)
tree.under <- rpart(DelayedArr~., data=undersampling)
tree.both <- rpart(DelayedArr~., data=bothsampling)

pred.tree.over <- predict(tree.over, newdata = dataset)
pred.tree.under <- predict(tree.under, newdata = dataset)
pred.tree.both <- predict(tree.both, newdata = dataset)
```
<br/>

* AUC Undersampling
```{r}
roc.curve(dataset$DelayedArr, pred.tree.under[,2])
```
<br/>

* AUC Oversampling
```{r}
roc.curve(dataset$DelayedArr, pred.tree.over[,2])
```
<br/>

* AUC Both
```{r}
roc.curve(dataset$DelayedArr, pred.tree.both[,2])
```

* Use either Undersampling or Bothsampling as it gives highest AUC score.

#### Logistic Regression
<br/>

* Train
```{r}
set.seed(1)
fit.glm_undersampling <- train(DelayedArr~., data=undersampling, method="glm", metric=metric, trControl=control, family="binomial")
set.seed(1)
fit.glm_oversampling <- train(DelayedArr~., data=oversampling, method="glm", metric=metric, trControl=control, family="binomial")
set.seed(1)
fit.glm_bothsampling <- train(DelayedArr~., data=bothsampling, method="glm", metric=metric, trControl=control, family="binomial")
```
<br/>

* Predict
```{r}
predictions_glm_undersampling <- predict(fit.glm_undersampling, validation)
predictions_glm_oversampling <- predict(fit.glm_oversampling, validation)
predictions_glm_bothsampling <- predict(fit.glm_bothsampling, validation)
```
<br/>

* Confusion Matrix
```{r}
confusionMatrix(predictions_glm_undersampling, validation$DelayedArr)
confusionMatrix(predictions_glm_oversampling, validation$DelayedArr)
confusionMatrix(predictions_glm_bothsampling, validation$DelayedArr)
```
<br/>

* Results
```{r}
results <- resamples(list(under=fit.glm_undersampling, over=fit.glm_oversampling, both=fit.glm_bothsampling))

summary(results)
```

#### Random Forest
<br/>

* Train
```{r}
set.seed(1)
fit.rf_undersampling <- train(DelayedArr~., data=undersampling, method="rf", metric=metric, trControl=control)
set.seed(1)
fit.rf_oversampling <- train(DelayedArr~., data=oversampling, method="rf", metric=metric, trControl=control)
set.seed(1)
fit.rf_bothsampling <- train(DelayedArr~., data=bothsampling, method="rf", metric=metric, trControl=control)
```
<br/>

* Predict
```{r}
predictions_rf_undersampling <- predict(fit.rf_undersampling, validation)
predictions_rf_oversampling <- predict(fit.rf_oversampling, validation)
predictions_rf_bothsampling <- predict(fit.rf_bothsampling, validation)
```
<br/>

* Confusion Matrix
```{r}
confusionMatrix(predictions_rf_undersampling, validation$DelayedArr)
confusionMatrix(predictions_rf_oversampling, validation$DelayedArr)
confusionMatrix(predictions_rf_bothsampling, validation$DelayedArr)
```
<br/>

* Results
```{r}
results <- resamples(list(under=fit.rf_undersampling, over=fit.rf_oversampling, both=fit.rf_bothsampling))

summary(results)
```

#### CART

<br/>

* Train
```{r}
set.seed(1)
fit.cart_undersampling <- train(DelayedArr~., data=undersampling, method="rpart", metric=metric, trControl=control)
set.seed(1)
fit.cart_oversampling <- train(DelayedArr~., data=oversampling, method="rpart", metric=metric, trControl=control)
set.seed(1)
fit.cart_bothsampling <- train(DelayedArr~., data=bothsampling, method="rpart", metric=metric, trControl=control)
```
<br/>

* Predict
```{r}
predictions_cart_undersampling <- predict(fit.cart_undersampling, validation)
predictions_cart_oversampling <- predict(fit.cart_oversampling, validation)
predictions_cart_bothsampling <- predict(fit.cart_bothsampling, validation)
```
<br/>

* Confusion Matrix
```{r}
confusionMatrix(predictions_cart_undersampling, validation$DelayedArr)
confusionMatrix(predictions_cart_oversampling, validation$DelayedArr)
confusionMatrix(predictions_cart_bothsampling, validation$DelayedArr)
```
<br/>

* Results
```{r}
results <- resamples(list(under=fit.cart_undersampling, over=fit.cart_oversampling, both=fit.cart_bothsampling))

summary(results)
```

#### KNN

<br/>

* Train
```{r}
set.seed(1)
fit.knn_undersampling <- train(DelayedArr~., data=undersampling, method="knn", metric=metric, trControl=control)
set.seed(1)
fit.knn_oversampling <- train(DelayedArr~., data=oversampling, method="knn", metric=metric, trControl=control)
set.seed(1)
fit.knn_bothsampling <- train(DelayedArr~., data=bothsampling, method="knn", metric=metric, trControl=control)
```
<br/>

* Predict
```{r}
predictions_knn_undersampling <- predict(fit.knn_undersampling, validation)
predictions_knn_oversampling <- predict(fit.knn_oversampling, validation)
predictions_knn_bothsampling <- predict(fit.knn_bothsampling, validation)
```
<br/>

* Confusion Matrix
```{r}
confusionMatrix(predictions_knn_undersampling, validation$DelayedArr)
confusionMatrix(predictions_knn_oversampling, validation$DelayedArr)
confusionMatrix(predictions_knn_bothsampling, validation$DelayedArr)
```
<br/>

* Results
```{r}
results <- resamples(list(under=fit.knn_undersampling, over=fit.knn_oversampling, both=fit.knn_bothsampling))

summary(results)
```
<br/>

* Total Result
```{r}
results_total <- resamples(list(under_glm=fit.glm_undersampling, over_glm=fit.glm_oversampling, both_glm=fit.glm_bothsampling,
under_rf=fit.rf_undersampling, over_rf=fit.rf_oversampling, both_rf=fit.rf_bothsampling,
under_cart=fit.cart_undersampling, over_cart=fit.cart_oversampling, both_cart=fit.cart_bothsampling,
under_knn=fit.knn_undersampling, over_knn=fit.knn_oversampling, both_knn=fit.knn_bothsampling))

summary(results_total)
```

* Among different models, we could pick the best model with appropriate value of accuracy, sensitivity and specificity which are a good indicators to evaluate the model.However, for this research, we will deep dive into Random Forest.

#### Tuning Random forest in R

* We will be using undersampling result to tune the model
```{r}
fit.rf_undersampling
```

* It shows that mtry of 5 has the highest accuracy for fit.rf_undersampling
<br/>

* Grid Search
```{r}
control_grid <- trainControl(method="cv", number=10 , search="grid")

set.seed(1)

fit.rf_undersampling_grid <- train(DelayedArr~., data=undersampling, method="rf", metric=metric, trControl=control_grid)

print(fit.rf_undersampling_grid)

plot(fit.rf_undersampling_grid)
```

* Grid-search takes quite long time to load, hence random search is often used instead. Although it is done within certain amount of time, it has systematic way to find the appropriate model instead of just putting random values several times. Moreover, it is also called coarse search because it proceeds two to three times and gradually narrows down.
<br/>

* RandomSearch
```{r}
control_random <- trainControl(method="cv", number=10, search="random")

set.seed(1)

fit.rf_undersampling_random <- train(DelayedArr~., data=undersampling, method="rf", metric=metric, trControl=control_random)

print(fit.rf_undersampling_random)

plot(fit.rf_undersampling_random)
```
<br/>

* Furthermore, to identify the importance of each variable, we will be using permutation based feature importance method to identify.

#### random forest model

```{r}
rf_permutation <- ranger(DelayedArr~ ., dataset, importance = 'permutation')
```
<br/>

* barplot
```{r}
rf_permutation_importance <- data.frame(
  var = names(rf_permutation$variable.importance), imp = c(rf_permutation$variable.importance))
rf_permutation_importance$data_type <- 'permutation'

ggplot(rf_permutation_importance) + 
  geom_bar(aes(x=reorder(var, imp), y=imp), stat = 'identity',fill="steelblue") +
  labs(title = "Importance of variables",x = "Variables", y ="Importance") + 
  theme(plot.title = element_text(hjust = 0.5))
```

* From the graph, we can identify that "CRSDepHour","Distance" and "Origin" have high influence to the model.Thus for further research, we could optimize the best parameter for get the best result of the model.
